\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-08-02 paper/hubble_paperclip.tex
update pdf

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training), a method which associates astronomical observations imaged by surveys and telescopes with natural language  using a neural network model. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful observing proposal abstracts, optionally summarized via guided generation using large language models (LLMs), and corresponding downstream observations. Using observations from the \hubble Space Telescope (HST) as an example, we show that the fine-tuned model embodies a meaningful joint representation between observations and text through tests targeting image retrieval (i.e., finding the most relevant observations using natural language queries) and description retrieval (i.e., querying for astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist \changes{foundation models} rather than task-specific models for \changes{interacting with} astronomical data by leveraging text as an interface. \githubmaster
\end{abstract}

@@

Machine learning (ML) is starting to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on augmenting traditional techniques in order to improve performance on specific tasks.
%
The {Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to simultaneously tackle a wide range of tasks \citep{bommasani2021opportunities}.
%
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised or weakly-supervised learning techniques, enabling them to learn powerful representations which can then be used downstream in different ways.
%
Foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the large amounts of publicly-available data and diverse modes of interacting with it.
%
The multi-modality inherent to astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given target object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to construct a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%

In this paper, we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training\footnote{Technically, we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and images corresponding to downstream observations. 
%
Concretely, we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract pairs results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
Our method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm.
%

The rest of this paper is organized as follows.
%
In Sec.~\ref{sec:dataset}, we describe the \hubble dataset used in this work, including the curation and processing of observations as well as text captions.
%
In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model.
%
In Sec.~\ref{sec:results}, we present quantitative and qualitative results of our experiments on retrieval tasks.
%
We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

\begin{figure*}[!t]
\centering
\includegraphics[width=0.99\textwidth]{plots/figure.pdf}
\caption{\changes{Overview of the PAPERCLIP method. (Left) A pre-trained CLIP model is fine-tuned using a dataset of \hubble observations and corresponding proposal abstracts. The proposal abstracts are optionally summarized using guided large language model generation. (Right) The fine-tuned model can then be used for downstream tasks such as observation retrieval (i.e., finding the observations most relevant to a given text query). The proposal abstract snippet shown here corresponds to proposal ID \href{https://archive.stsci.edu/proposal_search.php?id=16914&mission=hst}{16914}}.}
\label{fig:overview}
\end{figure*}

\section{\changes{Related work}}

% The CLIP family of foundation models, which in their original form embed images and associated captions into a common representation space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval~\citep{radford2021learning}.
%
The concept of learning task-agnostic representations via self-supervised and contrastive learning has been applied within astrophysics \citep{slijepcevic2024radio,stein2021self,hayat2021self,slijepcevic2022learning} and used for downstream tasks like object similarity search \citep{stein2021self}, gravitational lens finding \citep{stein2022mining}, estimation of Galactic distances \citep{hayat2021estimating}, identification of rare galaxies \citep{walmsley2023rare}, and data compression \citep{akhmetzhanova2024data}. For a recent review of contrastive learning in astrophysics, see \citet{huertas2023brief}. 
%

@@

% While the interaction of astrophysical data with large language models is fairly nascent, \textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is a recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

In this paper, we present for the first time an application associating target-agnostic astronomical data with the text modality, showing that this can be effectively accomplished through contrastive learning by leveraging observational proposals to inform text captions.
%

@@

Table \ref{tab:dataset} shows examples of images and their corresponding (clipped) proposal abstracts.
%
It can be seen that the images in this dataset exhibit specific characteristics as well as artifacts particular to HST data-taking and processing which distinguishes them from the distribution of natural images typically used for large-scale pre-training of foundation models \citep{deng2009imagenet}.
%
This further motivates the need for fine tuning on domain-specific data.

@@

\begin{table}[h!]
  \renewcommand{\arraystretch}{2}
    \centering
    \begin{tabular}{m{1.8cm} m{6.6cm} m{6.6cm}}
        \toprule
        \bfseries Prop. ID & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
        \cmidrule(r){2-3}
        & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
        \midrule
        \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
        \bottomrule
    \end{tabular}
    \caption{\changes{For the \hubble proposal abstracts shown in Tab.~\ref{tab:dataset},} the LLM (\textsc{Mixtral-8x7B})-extracted summaries showing objects and phenomena (middle column) as well as potential downstream science use cases (last column) separately.}
    \label{tab:datasetsumm}
\end{table}

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST)\footnote{\url{https://mast.stsci.edu/}} via the \package{Astroquery} \citep{2019AJ....157...98G} API.
%
Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images.
%
We note that these are not science-grade observations, but rather lower-resolution images useful for diagnostic purposes.
%
A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns.
%
Images are centered and resized to a resolution-per-side of 512 pixels.
%
Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on evaluation metrics.
%
If no appropriate images corresponding to an abstract are found, it is excluded from the dataset.

In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset.
%
3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of image-text pairs for evaluation. The held out images correspond to 438 unique abstracts.  % Double check numbers in final version.

@@

% Mention that the signal is noisy!

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the generation process by masking out tokens that would violate the intended format.
%
By framing the problem in terms of transitions between a set of finite states (a finite-state machine), \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation.
%
See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation method used here.

@@

The guided generation process ensures that, in this case, the output of the LLM strictly conforms to the format of the following example:
\begin{center}
  \begin{jsoncode}
    \centering
          \color{black}\{
            \color{codegreen}'objects_and_phenomena'\color{black}: [\color{codegreen}'star forming galaxy', 'lensed galaxy'\color{black}], 
            \color{codegreen}'science_use_cases'\color{black}: [\color{codegreen}'measure lensing magnification'\color{black}]
          \color{black}\}
  \end{jsoncode}
  \end{center}
which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
We train separate models using the raw abstracts and the LLM-generated summaries, and compare their performance on downstream tasks in Sec.~\ref{sec:results}.
%
We note that, even after summarization, the association signal is expected to be noisy, since parts of the summarized caption may not be directly descriptive of the observed images. The goal of the fine-tuning process is to leverage the signal contained in this noisy association.

@@

Our goal is to learn a semantically meaningful joint representation between images corresponding to HST observation and natural (English) language.
%
With PAPERCLIP, we leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

@@

Contrastive Language-Image Pre-training \citep[CLIP;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are considered on the same footing.

We use the \texttt{CLIP-ViT-B/16} \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension text sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$ pixels.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model contains $\sim 149$ million trainable parameters.
%
This model was originally pre-trained on 400 million image-text pairs from internet data \citep{radford2021learning}.
%

@@

All models were trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate $10^{-5}$ and weight decay $10^{-3}$.
%
Training takes approximately 3 hours on 4 Nvidia A100 GPUs.
Models were instantiated using the \package{Transformers} \citep{wolf2019huggingface} library and trained using packages from the \package{Jax} \citep{jax2018github} ecosystem.
%

@@

The model is evaluated by tracking the contrastive loss in Eq.~\eqrefb{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training.
%
The retrieval accuracy is defined as the fraction of associated captions which fall within the top $k\%$ of captions by cosine similarity of the normalized image and caption embeddings, averaged over the images in the validation set:
\begin{equation}
\text{Retrieval accuracy}_k = \frac{1}{|\mathcal V|} \sum_{i=1}^{|\mathcal V|} \mathbbm{1}\left[\operatorname{rank}\left({x}_i \cdot {y}_{i}; \{{x}_i \cdot {y}_{j}\}_{j=1}^{|\mathcal V|}\right) \leq \left\lfloor\frac{k}{100}|\mathcal V|\right\rfloor\right]
\label{eq:retrieval_accuracy}
\end{equation}
where $|\mathcal V|$ is the total number of images in the validation set, $\mathbbm{1}[\cdot]$ is the indicator function that returns 1 if the condition inside the brackets is true and 0 otherwise, $\operatorname{rank}\left({x}_i \cdot {y}_{i}; \{{x}_i \cdot {y}_{j}\}_{j=1}^{|\mathcal V|}\right)$ is a function that returns the rank of the cosine similarity between ${x}_i$ and ${y}_{i}$ among the cosine similarities between ${x}_i$ and all captions ${y}_j$ in the validation set, and $k$ is the percentage of top captions considered for the retrieval accuracy. Note that this metric is symmetric in the image and text modalities.

We also qualitatively evaluate the learned embeddings through image retrieval (i.e., retrieving the most relevant images from the validation set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation, akin to zero-shot classification) experiments. 
%
For the description/text retrieval evaluation, we define a list of possible text associations (i.e., classes), which we show in App.~\ref{app:categories}, by querying the \textsc{Claude 2}\footnote{\url{https://claude.ai/}} large language followed by with manual curation.

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy (right) evaluated on the held out validation set over the course of training, for different training configurations considered.
%
The dashed orange lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled.
%
This randomized baseline is seen to do on par with random expectation (i.e., a 10\% retrieval accuracy), unlike the others, validating the presence of a significant association signal between images and text in the dataset.
%
Interestingly, the base pre-trained model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$.
%
We therefore also compare the qualitative performance of the base model with the fine-tuned models on downstream retrieval tasks.

@@

Figure~\ref{fig:sim_valtrain} (right) shows the retrieval accuracy, as defined in Eq.~\eqrefb{eq:retrieval_accuracy}, as a function of the retrieval fraction $k\%$.
%
In this case, we evaluate all four models (fine tuned on raw abstracts (blue), fine-tuned on LLM-summarized abstracts (red), trained on LLM-summarized abstracts from scratch (yellow), and the base model (purple)) on the same captions dataset -- the summarized abstracts -- for a direct comparison.
%
Remarkably, the model trained on raw abstracts shows very similar performance when evaluated on the summarized abstracts compared to that trained on the summarized abstracts themselves, indicating that \emph{(1)} the image-text association signal is preserved in the summarization process, and \emph{(2)} the model is able to effectively leverage meaningful concepts in the noisy raw abstracts through weak supervision. \changes{The significantly worse performance of the model trained from scratch, compared to the fine-tuned models, highlights the crucial role of the inductive bias inherent in the base pretrained model, which effectively captures the rich associations between images and language.}

@@

\changes{We show retrieval accuracy performance for additional variations on the model and training configuration in App.~\ref{app:ablations}.}

@@

We list here the prompts and schema (i.e., desired output formats) used for guided text generation via \package{Outlines} package interfacing with the \textsc{Mixtral-8x7B-Instruct} open-weights LLM.

The following schema is used to guide the generation of the summaries, intended to produce between one and five objects and hypotheses, as well as science use cases. \\

@@

% \section{Additional Evaluation Metrics and Ablations}
% \label{app:ablations}

>>>>

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training), a method which associates astronomical observations imaged by telescopes with natural language using a neural network model. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful observing proposal abstracts and corresponding downstream observations, with the abstracts optionally summarized via guided generation using large language models (LLMs). Using observations from the \hubble Space Telescope (HST) as an example, we show that the fine-tuned model embodies a meaningful joint representation between observations and natural language through tests targeting image retrieval (i.e., finding the most relevant observations using natural language queries) and description retrieval (i.e., querying for astrophysical object classes and use cases most relevant to a given observation). Our study demonstrates the potential for using generalist foundation models rather than task-specific models for interacting with astronomical data by leveraging text as an interface. \githubmaster
\end{abstract}

@@

Machine learning (ML) is starting to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on augmenting traditional techniques in order to improve performance on specific tasks.
%
The \emph{foundation model} paradigm, in contrast, seeks to develop generalist models which can be deployed to simultaneously tackle a wide range of tasks \citep{bommasani2021opportunities}.
%
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised or weakly-supervised learning techniques, enabling them to learn powerful representations which can then be used downstream in different ways.
%
Foundation models can often benefit from additional training ({fine-tuning}) using a relatively small amount of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards,mccabe2023multiple,Birk:2024knn,vig2024finetuning,heinrich2024masked}, with astrophysics being ripe for such an effort given the large amounts of publicly-available data and diverse modes of interacting with it.
%
The multi-modality inherent to astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given target object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to construct a joint physically-informative embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%

In this paper, we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training\footnote{Technically, we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and images corresponding to downstream observations. 
%
Concretely, we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract pairs results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
Our method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm. A high-level overview of the method is shown in Fig.~\ref{fig:overview}.
%

The rest of this paper is organized as follows.
%
We review related work in Sec.~\ref{sec:related}.
%
In Sec.~\ref{sec:dataset}, we describe the \hubble dataset used in this work, including the curation and processing of observations as well as text captions.
%
In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model.
%
In Sec.~\ref{sec:results}, we present quantitative and qualitative results of our experiments on retrieval tasks.
%
We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

\begin{figure*}[!t]
\centering
\includegraphics[width=0.99\textwidth]{plots/figure.pdf}
\caption{Overview of the PAPERCLIP method. (Left) A pre-trained CLIP model is fine-tuned using a dataset of \hubble observations and corresponding proposal abstracts. The proposal abstracts are optionally summarized using guided large language model generation. (Right) The fine-tuned model can then be used for downstream tasks such as observation retrieval (i.e., finding the observations most relevant to a given text query). The proposal abstract snippet shown here corresponds to proposal ID \href{https://archive.stsci.edu/proposal_search.php?id=16914&mission=hst}{16914}.}
\label{fig:overview}
\end{figure*}

\section{Related Work}
\label{sec:related}

The concept of learning task-agnostic representations via self-supervised and contrastive learning has been applied within astrophysics \citep{slijepcevic2024radio,stein2021self,hayat2021self,slijepcevic2022learning} and used for downstream tasks like object similarity search \citep{stein2021self}, gravitational lens finding \citep{stein2022mining}, estimation of Galactic distances \citep{hayat2021estimating}, identification of rare galaxies \citep{walmsley2023rare}, and data compression \citep{akhmetzhanova2024data}. For a recent review of contrastive learning in astrophysics, see \citet{huertas2023brief}. 
%

@@

In this paper, we present for the first time an application associating target-agnostic astronomical data with the text modality, showing that this can be effectively accomplished through contrastive learning by leveraging observing proposal abstracts to inform text captions.
%

@@

Table \ref{tab:dataset} shows examples of images and their corresponding (clipped) proposal abstracts.
%
It can be seen that the images in this dataset exhibit specific characteristics as well as artifacts particular to HST data-taking and processing which distinguishes them from the distribution of natural images typically used for large-scale pre-training of foundation models.
%
This further motivates the need for fine-tuning on domain-specific data.

@@

\begin{table}[h!]
  \renewcommand{\arraystretch}{2}
    \centering
    \begin{tabular}{m{1.8cm} m{6.6cm} m{6.6cm}}
        \toprule
        \bfseries Prop. ID & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
        \cmidrule(r){2-3}
        & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
        \midrule
        \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
        \midrule
        \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
        \bottomrule
    \end{tabular}
    \caption{For the \hubble proposal abstracts shown in Tab.~\ref{tab:dataset}, the LLM (\textsc{Mixtral-8x7B})-extracted summaries showing objects and phenomena (middle column) as well as potential downstream science use cases (last column) separately. The proposal IDs (left column) contain hyperlinks to the MAST page corresponding to the proposal.}
    \label{tab:datasetsumm}
\end{table}

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST)\footnote{\url{https://mast.stsci.edu/}} via the \package{Astroquery} \citep{2019AJ....157...98G} API.
%
Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images.
%
We note that these are not science-grade observations, but rather lower-resolution images useful for diagnostic or preview purposes.
%
A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns.
%
Images are centered and resized to a resolution-per-side of 512 pixels.
%
Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on evaluation metrics.
%
If no appropriate images corresponding to an abstract are found, it is excluded from the dataset.

In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset.
%
3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of image-text pairs for evaluation. The held out images correspond to 429 unique abstracts.  % Double check numbers in final version.

@@

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the generation process by masking out tokens that would violate the intended format.
%
By framing the problem in terms of transitions between a set of finite states (i.e., a finite-state machine), \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation.
%
See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation method used here, including an overview of technical details.

@@

The guided generation process ensures that, in this case, the output of the generated output of the LLM strictly conforms to the format of the following example:\\
\begin{center}
  \begin{jsoncode}
    \centering
          \color{black}\{
            \color{deepgreen}'objects_and_phenomena'\color{black}: [\color{deepgreen}'star forming galaxy', 'lensed galaxy'\color{black}, ...], 
            \color{deepgreen}'science_use_cases'\color{black}: [\color{deepgreen}'measure lensing magnification'\color{black}, ...]
          \color{black}\}
  \end{jsoncode}
  \end{center}
which is then used to construct the summarized caption by combining the two key elements.
%
Examples of LLM-generated abstract summaries are shown in Tab.~\ref{tab:datasetsumm}, for the same set of abstracts as shown in Tab.~\ref{tab:dataset}.
%
We train separate models using the raw abstracts and the LLM-generated summaries, and compare their performance on downstream tasks in Sec.~\ref{sec:results}.
%
We note that, even after summarization, the association signal is expected to be noisy, since parts of the summarized caption may not be directly descriptive of the observed images. The goal of the fine-tuning process is to leverage the signal contained in this noisy association.

@@

Our goal is to learn a semantically meaningful joint representation between images corresponding to HST observation and natural (English) language.
%
With PAPERCLIP, we leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine-tuning.

@@

Contrastive Language-Image Pre-training \citep[CLIP;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text caption, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are considered on the same footing.

We use the \texttt{CLIP-ViT-B/16} \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer with patch size $16\times16$ as the image encoder \citep{dosovitskiy2020image} and a 12-layer, 8-head, 512-embedding dimension text sequence transformer as the text backbone \citep{vaswani2017attention}.
%
The text encoder has a maximum length of 77 tokens and the image encoder has a native resolution of $224\times224$ pixels.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model contains $\sim 149$ million trainable parameters.
%
This model was originally pre-trained on $\sim 400$ million image-text pairs from internet data.
%

@@

All models were trained over 20,000 steps with 2000 linear warmup steps 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate $10^{-5}$ and weight decay $10^{-3}$.
%
Training takes approximately 3 hours on 4 Nvidia A100 GPUs.
Models were instantiated using the \package{Transformers} \citep{wolf2019huggingface} library and trained using packages from the \package{Jax} \citep{jax2018github} ecosystem.
%

@@

The model is evaluated by tracking the contrastive loss in Eq.~\eqrefb{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training.
%
The retrieval accuracy is defined as the fraction of associated captions (either raw or LLM-summarized abstracts) which fall within the top $k\%$ of captions by cosine similarity of the normalized image and caption embeddings, averaged over the images in the validation set:
\begin{equation}
\text{Retrieval accuracy}_k = \frac{1}{|\mathcal V|} \sum_{i=1}^{|\mathcal V|} \mathbbm{1}\left[\operatorname{rank}\left({x}_i \cdot {y}_{i}; \{{x}_i \cdot {y}_{j}\}_{j=1}^{|\mathcal V|}\right) \leq \left\lfloor\frac{k}{100}|\mathcal V|\right\rfloor\right]
\label{eq:retrieval_accuracy}
\end{equation}
where $|\mathcal V|$ is the total number of images in the validation set, $\mathbbm{1}[\cdot]$ is the indicator function that returns 1 if the condition inside the brackets is true and 0 otherwise, $\operatorname{rank}\left({x}_i \cdot {y}_{i}; \{{x}_i \cdot {y}_{j}\}_{j=1}^{|\mathcal V|}\right)$ is a function that returns the rank of the cosine similarity between ${x}_i$ and ${y}_{i}$ among the cosine similarities between ${x}_i$ and all captions ${y}_j$ in the validation set, and $k$ is the percentage of top captions considered for the retrieval accuracy. Note that this metric is symmetric in the image and text modalities.

We also qualitatively evaluate the learned embeddings through image retrieval (i.e., retrieving the most relevant images from the validation set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation, akin to zero-shot classification) experiments. 
%
For the description/text retrieval evaluation, we define a list of possible text associations (i.e., classes), which we show in App.~\ref{app:categories}, by querying the \textsc{Claude 2}\footnote{\url{https://claude.ai/}} large language followed by manual curation.

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy (right) evaluated on the held out validation set over the course of training, for different training configurations considered.
%
The dashed orange lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled.
%
This randomized baseline is seen to do on par with random expectation (i.e., a 10\% retrieval accuracy), unlike the others, validating the presence of a significant association signal between images and text in the dataset.
%
Interestingly, the base pre-trained model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$ (as see from the left-most datum in Fig.~\ref{eq:retrieval_accuracy} right, for the curves corresponding to fine-tuned models).
%
We therefore also compare the qualitative performance of the base model with the fine-tuned models on downstream retrieval tasks.

@@

Figure~\ref{fig:sim_valtrain} (right) shows the retrieval accuracy, as defined in Eq.~\eqrefb{eq:retrieval_accuracy}, as a function of the retrieval fraction $k\%$.
%
In this case, we evaluate all four models (fine-tuned on raw abstracts (blue), fine-tuned on LLM-summarized abstracts (red), trained on LLM-summarized abstracts from scratch (yellow), and the base model (purple)) on the same captions dataset -- the summarized abstracts -- for a direct comparison.
%
Remarkably, the model trained on raw abstracts shows very similar performance when evaluated on the summarized abstracts compared to that trained on the summarized abstracts themselves, indicating that \emph{(1)} the image-text association signal is preserved in the summarization process, and \emph{(2)} the model is able to effectively leverage meaningful concepts in the noisy raw abstracts through weak supervision. The significantly worse performance of the model trained from scratch, compared to the fine-tuned models, highlights the crucial role of the inductive bias inherited from the base pre-trained model, which effectively captures rich associations between images and language.

@@

We show retrieval accuracy performance for additional variations on the model and training configuration in App.~\ref{app:ablations}.

@@

We list here the prompts and schema (i.e., desired output formats) used for guided text generation via \package{Outlines} package interfacing with the \textsc{Mixtral-8x7B-Instruct} open-weights large language model.

The following schema, specified using the data-validation package \package{Pydantic}, is used to guide the generation of the summaries, intended to produce between one and five objects and hypotheses, as well as science use cases, given a raw proposal abstract. Both fields are of type \texttt{conlist}, a \package{Pydantic} type that represents a constrained list.  \\

@@
